A quick user guide to the\\ \scine-\qcm\ software suite for \mol

### 2021-08-24 dmrg/doc/qcmaquis_manual.tex

\subsection{QCMaquis input example}\label{sec:input-qcm}

@@

\begin{lstlisting}[language=qcmaquis,caption={Input example for a DMRG-CASCI(8,8) standalone \qcm{} calculation},label=lst:qcm]
 //active-space dependent parameters
L     = 8
nelec = 8
// target symmetry (spin and spatial) of the wave function:
// 2*spin = 0 (singlet) + totally symmetric point group irrep
spin  = 0
irrep = 0
//parameters to control the actual DMRG calculation
nsweeps = 8
max_bond_dimension = 256
conv_thresh = 1e-6
// initialization procedure
init_state = 'default'
// technical parameters
symmetry = 'su2u1pg'
integral_cutoff = 1e-40
truncation_initial = 1e-50
truncation_final = 1e-7
chkpfile = 'chkp.h5'
resultfile = 'result.h5'
integral_file = "FCIDUMP"
storagedir = '/scratch/$USER/boundaries'
LATTICE = "orbitals"
lattice_library = "coded"
MODEL = "quantum_chemistry"
model_library = "coded"
//all expectation value calculations required for entropy measures
MEASURE[ChemEntropy]
\end{lstlisting}

@@

The pre-Born--Oppenheimer (pre-BO) module within QCMaquis provides a tool to perform
DMRG-CASCI calculations of \textit{multicomponent} systems consisting of multiple spin-$\frac12$
particle types.

QCMaquis follows the formatting of the integral file shown in
Listing~\ref{lst:prebodump}. The first line with only a single float,
contains a constant shift in energy, e.g., point-charge repulsion of classical
nuclei.
The following lines contain the terms in the Hamiltonian where pairs of numbers
$m-n$ describe a single second-quantization operator. The first number denotes
the particle type (starting with 0) and the second number the orbital index of
the given particle type (also starting with 0 for all types). 
We assume physics notation and normal-ordering, i.e., the following would be a
valid term
\begin{equation}
  \left(\int \phi_{1-1}(1)\phi_{0-1}(2)g(1,2)\phi_{1-0}(1)\phi_{0-1}(2)\right)
  a^\dagger_{1-1} a^\dagger_{0-1} a_{0-1} a_{1-0}\,.
\end{equation}
The second and third lines contain single-paricle type \textit{hopping} terms.
The fourth line stands for a single-particle type interaction, and the fifth
line for an interaction between different particle types. 
Note here that the aforementioned convention implies that operators belonging to
the same particle types in a given string of operators must be either at
positions 1 and 4 or 2 and 3.

@@

\qcm\ requires only two files to start a calculation:
\begin{enumerate}
 \item an input file, see Listing~\ref{lst:qcm} for example
 \item an integral file in the \texttt{FCIDUMP} format as described in Ref. \citenum{fcidump}.
\end{enumerate}

@@

\subsection{Typical workflow for a standalone QCMaquis DMRG calculation}\label{sec:workflow-qcm}
The usual workflow to set up, run and analyze a DMRG calculation proceeds as follows:
\begin{itemize}
   \item prepare an FCIDUMP integral file using the \mol-\qcm\ host
   program driver \cite{interface}) starting from a set of previously computed reference orbitals
   \item prepare an input file \texttt{dmrg-input} starting for example from the template input file provided in Listing~\ref{lst:qcm} and adjust all molecular system and wave function specific parameters (for example \texttt{nelec, spin,
   irrep, L, \ldots}, see Sections \ref{sec:compul-qcm}-\ref{sec:prop-qcm} for a list of all compulsory and optional input arguments)
   \item run the DMRG calculation with\\
\texttt{dmrg dmrg-input (| tee out)}
   \item compute expectation values with\\
\texttt{dmrg\_meas dmrg-input}
   \item analyze the results using the \textsc{Python} tools provided with the \qcm\ package (see Sections \ref{sec:tools-qcm}\ and
   \ref{sec:pytools-qcm}, respectively, for a list of utility programs and scripts)
\end{itemize}

>>>>

\subsection{QCMaquis input example}
\label{sec:input-qcm}

@@

\begin{lstlisting}[language=qcmaquis,caption={Input example for a DMRG-CASCI(8,8) standalone \qcm{} calculation},label=lst:qcm]
  // Lattice size (= number of orbitals)
  L     = 8
  // Number of electrons
  nelec = 8
  // Target symmetry (spin and spatial) of the wave function:
  // 2*spin = 0 (singlet)
  spin  = 0
  // Wavefunction irreducible representation (0 == totally symmetric)
  irrep = 0
  // Parameters to control the actual DMRG calculation
  nsweeps              = 8
  max_bond_dimension   = 256
  conv_thresh          = 1e-6
  // MPS initialization
  init_state = 'default'
  // Wave function symmetry (su2u1pg for spin-adapted DMRG + point group)
  symmetry             = 'su2u1pg'
  // Threshold below which integrals inside the integral_file are neglected
  integral_cutoff      = 1e-40
  // Threshold for the SVD truncation for the initial sweeps
  truncation_initial   = 1e-16
  // Threshold for the SVD truncation for the final sweeps
  truncation_final     = 1e-7
  // HDF5 file where the MPS will be stored
  chkpfile             = 'chkp.h5'
  // HDF5 file where the results of the DMRG calculations will be stored
  resultfile           = 'result.h5'
  // File with the one- and two-body integrals
  integral_file        = "FCIDUMP"
  // Path to the scratch folder
  storagedir           = '/scratch/$USER/boundaries'
  // Lattice-related parameters (not to be changed for qc calculations)
  LATTICE              = "orbitals"
  lattice_library      = "coded"
  MODEL                = "quantum_chemistry"
  model_library        = "coded"
  // List of measurements to be performed at the end of the optimization
  MEASURE[ChemEntropy] = 1
\end{lstlisting}

\subsection{\qcm\ examples library}
\label{sec:examples}

We provide in the folder \texttt{example} a set of example input files to run the different kinds of DMRG-based calculations that are supported by \qcm{}.
These example calculations are organized in different folder, one per feature provided by \qcm, as listed below:

\begin{itemize}
 \item \texttt{iTD-DMRG}: example calculation (\texttt{H2\_2e4o.iTD.SS.inp}) of an imaginary-time DMRG calculation. The integral file is generated for the H$_2$ molecule and based on the 6-31G basis set.
 Note that the calculation should be run with the executable \texttt{applications/evolve/dmrg\_evolve}, which is generated by setting the \texttt{BUILD\_DMRG\_EVOLVE} CMake flag to \texttt{ON}.
 We provide also the input for running a conventional \texttt{DMRG} calculation (\texttt{H2\_2e4o.TI.SS.inp}), which must be run with the \texttt{applications/dmrg/dmrg} executable and should return the same energy as the iTD-DMRG simulation.
 \item \texttt{TD-DMRG}: example calculation of a real-time DMRG calculation. The integral file is associated with the $\pi$ orbital space (CAS(6,6)) of benzene calculated with the 6-31G* basis set. As already noted above, the calculations should be run with the \texttt{applications/evolve/dmrg\_evolve} executable.
 Note that the initial state for the propagation is a random MPS (the \texttt{init\_state} keyword is set to \texttt{default}).
 Therefore, the calculation does not correspond to any physically meaningful non-equilibrium process. 
 However, this calculation can be used to assess that the \texttt{dmrg\_evolve} executable has been compiled correctly by verifying that the energy is conserved along the propagation.
 For this example calculation we provide the input files for both the single-site (\texttt{Benzene\_6e6o.TD.SS.inp}) and the two-site (\texttt{Benzene\_6e6o.TD.TS.inp}) propagators.
\end{itemize}

For all input files, we provide also the corresponding output file that can be used as a reference to ensure that the \qcm\ executables have been compiled correctly.

@@

The \qcm\ pre-Born--Oppenheimer (pre-BO) module provides a tool to perform DMRG-CASCI calculations for \textit{multicomponent} Hamiltonians consisting of multiple spin-$\frac12$ particle types.

QCMaquis follows the formatting of the integral file shown in Listing~\ref{lst:prebodump}.
The first line, with only a single float, contains a constant shift in energy, e.g., point-charge repulsion of classical nuclei.
The following lines contain the terms in the Hamiltonian where pairs of numbers $m-n$ describe a single second-quantization operator.
The first number denotes the particle type (starting with 0) and the second number the orbital index associated with the given particle type (also starting with 0 for all types).
We assume physics notation and normal-ordering, i.e., the following would be a valid term

\begin{equation}
  \left(\int \phi_{1-1}(1)\phi_{0-1}(2)g(1,2)\phi_{1-0}(1)\phi_{0-1}(2)\right)
  a^\dagger_{1-1} a^\dagger_{0-1} a_{0-1} a_{1-0}\,.
\end{equation}

The second and third lines contain single-paricle type \textit{hopping} terms.
The fourth line stands for a single-particle type interaction, and the fifth
line for an interaction between different particle types. 
Note here that the aforementioned convention implies that operators belonging to
the same particle types in a given string of operators must be either at
positions 1 and 4 or 2 and 3.

@@

\qcm\ requires only two files to start a calculation:

\begin{enumerate}
 \item an input file, see Listing~\ref{lst:qcm} for example
 \item an integral file in the \texttt{FCIDUMP} format as described in Ref. \citenum{fcidump}.
\end{enumerate}

@@

\subsection{Typical workflow for a standalone QCMaquis DMRG calculation}
\label{sec:workflow-qcm}

The usual workflow to set up, run and analyze a DMRG calculation proceeds as follows:

\begin{itemize}
   \item prepare an FCIDUMP integral file using the \mol-\qcm\ host
   program driver \cite{interface}) starting from a set of previously computed reference orbitals
   \item prepare an input file \texttt{dmrg-input} starting for example from the template input file provided in Listing~\ref{lst:qcm} and adjust all molecular system and wave function specific parameters (for example \texttt{nelec, spin,
   irrep, L, \ldots}, see Sections \ref{sec:compul-qcm}-\ref{sec:prop-qcm} for a list of all compulsory and optional input arguments)
   \item run the DMRG calculation with\\
\texttt{dmrg dmrg-input (| tee out)}
   \item compute expectation values with\\
\texttt{dmrg\_meas dmrg-input}
   \item analyze the results using the \textsc{Python} tools provided with the \qcm\ package (see Sections \ref{sec:tools-qcm}\ and
   \ref{sec:pytools-qcm}, respectively, for a list of utility programs and scripts)
\end{itemize}
